---
title: 'The cumulative distribution function'
teaching: 3
exercises: 0
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is the empirical cumulative distribution function?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Demonstrate the cumulative distribution function with an example.

::::::::::::::::::::::::::::::::::::::::::::::::


```{r message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
set.seed(51) # set a seed for reproducibility
frog_counts <-rpois(n = 200, lambda = 4)
frog_sizes <- rnorm(n = 200, mean = 7, sd = 2)
frog_counts_different_lakes <- rnbinom(n=200, size=2, mu=4)  

```



The cumulative distribution is the integral of the distribution, and thus the empirical cumulative distribution is the integral of the histogram. It gives you the percentage of values that are smaller than a certain value.  
We can visualize it with `stat_ecdf` in `ggplot2`.

```{r ecdf-plot, echo=TRUE, fig.height=3.5}
data.frame(frog_sizes) %>% 
  ggplot(aes(x=frog_sizes))+
  stat_ecdf()
```

As for the histogram, you could calculate the expected values for the cumulative distribution and add them to the plot:

```{r ecdf-cdf-plot, echo=TRUE, fig.height=3.5}
sizes <- seq(0,14,.5)
theoretical_cdf <- pnorm(sizes, mean=7, sd=2)

data.frame(frog_sizes) %>% 
  ggplot(aes(frog_sizes)) +
  stat_ecdf()+
  geom_line(data=data.frame(sizes,theoretical_cdf), aes(sizes, theoretical_cdf),colour="green")
```

Off this plot you can read for a frog size of $x$ (e.g. 7cm), that about 50% of the values are smaller than this value. So this is really just another way to display the histogram.  

:::::::::::::::::::::: callout
## Note
The procedure above was shown for demonstration purposes. We will not practice overlaying empirical and theoretical distribution functions with `ggplot`, because in the next episode, we'll introduce a tool that does the work for you.
:::::::::::::::::::::::::::

